---
title: "Determining Risk Factors for Injuries and Fatalities in Road Collisions"
authors: "Reuben Walker, Keegan MacDonald, Sanita Richards & Corey Thorne"
output:
  html_document:
    df_print: paged
---

## Introduction
The Canadian National Collision Database contains information on all
police-reported motor vehicle collisions over many years. The attached dataset
"2019_dataset_en.csv" contains such information for 272301 collisions in the
year 2019.

This R Notebook will load this dataset, clean, inspect and visualize its
contents, and produce a machine learning model to predict the severity of a
collision (whether it resulted in no injuries, non-fatal, or fatal injuries)
given a subset of the other available information.

## Loading libraries
The following libraries will be required to run this R Notebook:

```{r}
library(readr)
library(dplyr)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library(GGally)
library(data.table)
library(sf)
library(ggspatial)
library(tmap)
library(grid)
library(gridExtra)
library(patchwork)
```

## Loading the dataset
The dataset will be loaded as a CSV into the active R environment, it must be
located in the same directory as this R Notebook:

```{r}
data <- read_csv("2019_dataset_en.csv", show_col_types = FALSE)
```

## Cleaning the data
To clean the data, six columns containing unique identifiers, the year (2019),
and other variables not relevant or inhibitory to this machine learning task
must be removed as follows:

```{r}
data <- data %>%
  select(-c(C_YEAR, C_SEV, V_ID, P_ID, P_SAFE, C_CASE))
```

For the meanings of these column names, refer to the attached
"data_dictionary.pdf" for details.

The remaining columns are almost exclusively categorical variables, and contain
categories for unknown and missing data. These must be removed for the machine
learning model to process the dataset.

```{r}
data <- data %>%
  filter(C_MNTH!="UU",C_MNTH!="XX") %>%
  filter(C_WDAY!="U") %>% 
  filter(C_HOUR!="UU") %>%
  filter(C_VEHS!="UU",C_VEHS!="XX") %>%
  filter(C_CONF!="QQ",C_CONF!="UU",C_CONF!="XX") %>%
  filter(C_RCFG!="QQ",C_RCFG!="UU",C_RCFG!="XX") %>% 
  filter(C_WTHR!="U",C_WTHR!="X",C_WTHR!="Q") %>% 
  filter(C_RSUR!="U",C_RSUR!="X",C_RSUR!="Q") %>%
  filter(C_RALN!="U",C_RALN!="X",C_RALN!="Q") %>%
  filter(C_TRAF!="QQ",C_TRAF!="UU",C_TRAF!="XX") %>%
  filter(V_TYPE!="NN",V_TYPE!="QQ",V_TYPE!="UU",V_TYPE!="XX") %>%
  filter(V_YEAR!="NNNN",V_YEAR!="UUUU",V_YEAR!="XXXX") %>%
  filter(P_SEX!="N",P_SEX!="U",P_SEX!="X") %>%
  filter(P_AGE!="NN",P_AGE!="UU",P_AGE!="XX") %>%
  filter(P_PSN!="NN",P_PSN!="QQ",P_PSN!="UU",P_PSN!="XX") %>%
  filter(P_USER!="U") %>%
  filter(P_ISEV!="N",P_ISEV!="U")
```

Removing these leaves 17 of the original 23 columns and 167803 of the original
272301 observations. Finally, the most of remaining data must be converted to
factors / categorical variables, and some variables will be grouped into fewer
categories than they began with.

```{r}
# Group the C_CONF column while naming categories.
data <- data %>%
  mutate(C_CONF = (ifelse(data$C_CONF <= 10, 'Single Vehicle',
                    ifelse(data$C_CONF>20 & data$C_CONF <30, '2 Vehicles',
                    ifelse(data$C_CONF>30 & data$C_CONF<40,'2 Vehicles Head On','2 Vehicles Parked')))))

# Convert most variables to factors and some to numeric.
data$C_MNTH <- as.factor(data$C_MNTH)
data$C_WDAY <- as.factor(data$C_WDAY)
data$C_HOUR <- as.numeric(data$C_HOUR)
data$C_VEHS <- as.numeric(data$C_VEHS)
data$C_CONF <- as.factor(data$C_CONF)
data$C_RCFG <- as.factor(data$C_RCFG)
data$C_WTHR <- as.factor(data$C_WTHR)
data$C_RSUR <- as.factor(data$C_RSUR)
data$C_RALN <- as.factor(data$C_RALN)
data$C_TRAF <- as.factor(data$C_TRAF)
data$V_TYPE <- as.factor(data$V_TYPE)
data$V_YEAR <- as.numeric(data$V_YEAR)
data$P_SEX <- as.factor(data$P_SEX)
data$P_AGE <- as.numeric(data$P_AGE)
data$P_PSN <- as.factor(data$P_PSN)
data$P_ISEV <- as.factor(data$P_ISEV)
data$P_USER <- as.factor(data$P_USER)
```

To improve the readability of the data for the remainder of this script, the
values of categorical variables and the columns will be assigned English names:

```{r}
# Assign names to values of categorical variables.
data <- data %>%
  mutate(C_MNTH = recode(C_MNTH,
                        "1" = "Jan", "2" = "Feb", "3" = "Mar", "4" = "Apr",
                        "5" = "May", "6" = "Jun", "7" = "Jul", "8" = "Aug",
                        "9" = "Sep", "10" = "Oct", "11" = "Nov", "12" = "Dec")) %>%
  mutate(C_WDAY = recode(C_WDAY,
                        "1" = "Mon", "2" = "Tue", "3" = "Wed", "4" = "Thu",
                        "5" = "Fri", "6" = "Sat", "7" = "Sun")) %>%
  mutate(C_RCFG = recode(C_RCFG,
                         "1" = "Non-Intersection",
                         "2" = "Intersection - 2 public roadways",
                         "3" = "Intersection - parking, driveway etc.",
                         "4" = "Railroad level crossing",
                         "5" = "Bridge, overpass, viaduct",
                         "6" = "Tunnel or underpass",
                         "7" = "Passing or climbing lane",
                         "8" = "Ramp",
                         "9" = "Traffic circle",
                         "10" = "Express lane of a freeway",
                         "11" = "Collector lane of a freeway",
                         "12" = "Transfer lane of a freeway")) %>%
  mutate(C_WTHR = recode(C_WTHR,
                         "1" = "Clear & sunny", "2" = "Cloudy, no precipitation",
                         "3" = "Raining", "4" = "Snowing",
                         "5" = "Freezing rain, sleet, hail",
                         "6" = "Visibility limitation", "7" = "Strong wind")) %>%
  mutate(C_RSUR = recode(C_RSUR,
                         "1" = "Dry, normal", "2" = "Wet", "3" = "Snow",
                         "4" = "Slush, wet snow", "5" = "Icy", "6" = "Sand/gravel/dirt",
                         "7" = "Muddy", "8" = "Oil", "9" = "Flooded")) %>%
  mutate(C_RALN = recode(C_RALN,
                         "1" = "Straight and level", "2" = "Straight with gradient",
                         "3" = "Curved and level", "4" = "Curved with gradient",
                         "5" = "Top of hill or gradient", "6" = "bottom of hill or gradient")) %>%
  mutate(C_TRAF = recode(C_TRAF,
                         "1" = "Traffic signals fully operational",
                         "2" = "Traffic signals in flashing mode", "3" = "Stop sign",
                         "4" = "Yield sign", "5" = "Warning sign",
                         "6" = "Pedestrian crosswalk", "7" = "Police officer",
                         "8" = "School guard, flagman", "9" = "School crossing",
                         "10" = "Reduced speed zone", "11" = "No passing zone sign",
                         "12" = "Marking on the road",
                         "13" = "School bus - signal lights flashing",
                         "14" = "School bus - signal lights not flashing",
                         "15" = "Railway crossing with signals, gates",
                         "16" = "Railway crossing with signs only",
                         "17" = "Control device not specified",
                         "18" = "No control present")) %>%
  mutate(V_TYPE = recode(V_TYPE,
                         "1" = "Light duty vehicle", "5" = "Panel/cargo van",
                         "6" = "Other trucks and vans", "7" = "Unit trucks > 4536 kg",
                         "8" = "Road tractor", "9" = "School bus",
                         "10" = "Smaller school bus", "11" = "Urban and intercity bus",
                         "14" = "Motorcycle and moped", "16" = "Off road vehicles",
                         "17" = "Bicycle", "18" = "Purpose-built motorhome",
                         "19" = "Farm equipment", "20" = "Construction equipment",
                         "21" = "Fire engine", "22" = "Snowmobile", "23" = "Streetcar")) %>%
  mutate(P_PSN = recode(P_PSN,
                        "11" = "Driver", "12" = "Front row, center",
                        "13" = "Front row, right outboard", "21" = "Second row, left outboard",
                        "22" = "Second row, center", "23" = "Second Row, right outboard",
                        "31" = "Third row, left outboard", "32" = "Third row, center",
                        "33" = "Third row, right outboard", "96" = "Position unknown, but on-board",
                        "97" = "Sitting on someone's lap", "98" = "Outside passenger compartment",
                        "99" = "Pedestrian")) %>%
  mutate(P_ISEV = recode(P_ISEV,
                         "1" = "No injury", "2" = "Injury", "3" = "Fatality")) %>%
  mutate(P_USER = recode(P_USER,
                         "1" = "Motor vehicle driver", "2" = "Motor vehicle passenger",
                         "3" = "Pedestrian", "4" = "Bicyclist", "5" = "Motorcyclist"))
# Assign column names.
names(data) <- c("Month", "Weekday", "Hour", "Vehicles", "CollisionType", 
                   "RoadConfig", "Weather", "RoadSurface", 
                  "RoadAlignment", "Traffic", "VehicleType", 
                  "VehicleYear", "PersonSex", "PersonAge", 
                  "PersonPosition", "InjurySeverity", "UserType")

```

With these modifications, the data is mostly prepared for machine learning tasks,
and must now be visualized to show any more subtle problems that may appear
during training of a model.

## Data visualization
```{r}
# Set up histogram colors for histograms
InjurySeverity.colors <- c('Fatality' = 'brown1', 'Injury' = 'darkgoldenrod1', 'No injury' = 'darkseagreen4')

# This bit of code ensures that months display in the proper order on the bar charts
data$Month <- factor(data$Month, levels=c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec'))

#Create initial histograms for EDA:

#Month
monthHist <- ggplot(data, aes(x=Month,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

monthHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Month,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

#Weekday
weekdayHist <- ggplot(data, aes(x=Weekday,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

weekdayHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Weekday,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

#Hour
hourHist <- ggplot(data, aes(x=Hour,fill=InjurySeverity))+
  geom_bar(bins=24)+
  scale_fill_manual(values=InjurySeverity.colors)

hourHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Hour,fill=InjurySeverity))+
  geom_bar(bins=24)+
  scale_fill_manual(values=InjurySeverity.colors)

#Number of vehicles involved
numVehiclesHist <- ggplot(data, aes(x=Vehicles,fill=InjurySeverity))+
  geom_bar(bins=24)+
  scale_fill_manual(values=InjurySeverity.colors)

numVehiclesHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Vehicles,fill=InjurySeverity))+
  geom_bar(bins=24)+
  scale_fill_manual(values=InjurySeverity.colors)

#RoadConfig
roadConfigHist <- ggplot(data, aes(x=RoadConfig,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

roadConfigHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=RoadConfig,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#Weather
weatherHist <- ggplot(data, aes(x=Weather,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

weatherHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Weather,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#RoadSurface
roadSurfaceHist <- ggplot(data, aes(x=RoadSurface,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

roadSurfaceHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=RoadSurface,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#RoadAlignment
roadAlignmentHist <- ggplot(data, aes(x=RoadAlignment,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

roadAlignmentHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=RoadAlignment,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#Traffic
trafficHist <- ggplot(data, aes(x=Traffic,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

trafficHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Traffic,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#VehicleType
vehTypeHist <- ggplot(data, aes(x=VehicleType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

vehTypeHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=VehicleType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#VehicleYear
vehYearHist <- ggplot(data, aes(x=VehicleYear,fill=InjurySeverity))+
  geom_bar(bins=100)+
  scale_fill_manual(values=InjurySeverity.colors)

vehYearHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=VehicleYear,fill=InjurySeverity))+
  geom_bar(bins=100)+
  scale_fill_manual(values=InjurySeverity.colors)

#PersonSex
sexHist <- ggplot(data, aes(x=PersonSex,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

sexHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=PersonSex,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

#PersonAge
ageHist <- ggplot(data, aes(x=PersonAge,fill=InjurySeverity))+
  geom_bar(bins=99)+
  scale_fill_manual(values=InjurySeverity.colors)

ageHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=PersonAge,fill=InjurySeverity))+
  geom_bar(bins=96)+
  scale_fill_manual(values=InjurySeverity.colors)

#PersonPosition
pPositionHist <- ggplot(data, aes(x=PersonPosition,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

pPositionHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=PersonPosition,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#UserType
pUserTypeHist <- ggplot(data, aes(x=UserType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

pUserTypeHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=UserType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

#collisionType
collisionTypeHist <- ggplot(data, aes(x=CollisionType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

collisionTypeHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=CollisionType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)
```

First, we create bar charts to see how many collisions of each severity there are according to each variable. Most of the variables are either categorical, or integer numbers so bar charts are the best option.

Bar charts are displayed grouped by theme:

 - Time
 - Demographics
 - Weather & road conditions
 - Vehicle info
 - Road info
 - Driver/passenger info
 - Collision info
 
 Note that for all variables, the amount of fatalities is drastically lower than the amount of injuries and non-injuries, to the point where it's impossible to view patterns in the data. Therefore, for each varaible there is a separate chart to show how fatalities correlate with each variable.

```{r, out.width="100%"}
#Display similar grouped histograms:

#Histograms for time/day/month
timeHists <- (monthHist|weekdayHist|hourHist) & theme(legend.position = "right")
timeHistsFatal <- (monthHistFatal|weekdayHistFatal|hourHistFatal) & theme(legend.position = "none")
(timeHists/timeHistsFatal) + plot_layout(guides = "collect")

#Histograms for demographics
demoHists <- sexHist + ageHist & theme(legend.position = "right")
demoHistsFatal <- sexHistFatal + ageHistFatal & theme(legend.position = "none")
(demoHists/demoHistsFatal) + plot_layout(guides = "collect")

#Histograms for road & weather conditions
weatherRoadCondHists <- weatherHist + roadSurfaceHist & theme(legend.position = "right")
weatherRoadCondHistsFatal <- weatherHistFatal + roadSurfaceHistFatal & theme(legend.position = "none")
(weatherRoadCondHists/weatherRoadCondHistsFatal) + plot_layout(guides = "collect")

#Histograms for vehicle info
vehicleInfoHists <- vehTypeHist + vehYearHist & theme(legend.position = "right")
vehicleInfoHistsFatal <- vehTypeHistFatal + vehYearHistFatal & theme(legend.position = "none")
(vehicleInfoHists/vehicleInfoHistsFatal) + plot_layout(guides = "collect")

#Histograms for road info
roadInfoHists <- roadAlignmentHist + trafficHist + roadConfigHist & theme(legend.position = "right")
roadInfoHistsFatal <- roadAlignmentHistFatal + trafficHistFatal + roadConfigHistFatal & theme(legend.position = "none")
(roadInfoHists/roadInfoHistsFatal) + plot_layout(guides = "collect")

#Histograms for driver/passenger info
driverPassengerHists <- pPositionHist + pUserTypeHist & theme(legend.position = "right")
driverPassengerHistsFatal <- pPositionHistFatal + pUserTypeHistFatal & theme(legend.position = "none")
(driverPassengerHists/driverPassengerHistsFatal) + plot_layout(guides = "collect")

#Histograms for collision info
collisionHists <- collisionTypeHist + numVehiclesHist & theme(legend.position = "right")
collisionHistsFatal <- collisionTypeHistFatal + numVehiclesHistFatal & theme(legend.position = "none")
(collisionHists/collisionHistsFatal) + plot_layout(guides = "collect")
```


The data is distributed in such a way that the majority of the results of the target variable 
InjurySeverity are either No injury or Injury. No injury makes up (76530)/167803 or 45.6% of cases, 
Injury makes up (90299)/167807 or 53.8% leaving only 974/167807 or 0.5% resulting in fatalities.  
Fortunately, given the size of the data, we can acquire 3 suitably equal sized samples of the data 
for training and testing purposes.


```{r}
fatalSet <- data %>%
  filter(InjurySeverity == "Fatality") %>%
  sample_n(900)

injurySet <- data %>%
  filter(InjurySeverity == "Injury") %>%
  sample_n(900)

noInjurySet <- data %>%
  filter(InjurySeverity == "No injury") %>%
  sample_n((900))

data <- bind_rows(fatalSet,injurySet,noInjurySet, .id = NULL)

```




