---
title: "Determining Risk Factors for Injuries and Fatalities in Road Collisions"
authors: "Reuben Walker, Keegan MacDonald, Sanita Richards & Corey Thorne"
output:
  html_document:
    df_print: paged
---

## Introduction
The Canadian National Collision Database contains information on all
police-reported motor vehicle collisions over many years. The attached dataset
"2019_dataset_en.csv" contains such information for 272301 collisions in the
year 2019.

This R Notebook will load this dataset, clean, inspect and visualize its
contents, and produce a machine learning model to predict the severity of a
collision (whether it resulted in no injuries, non-fatal, or fatal injuries)
given a subset of the other available information.

## Loading libraries
The following libraries will be required to run this R Notebook:

```{r}
library(readr)
library(dplyr)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library(GGally)
library(data.table)
library(sf)
library(ggspatial)
library(tmap)
library(grid)
library(gridExtra)
library(patchwork)
```

## Loading the dataset
The dataset will be loaded as a CSV into the active R environment, it must be
located in the same directory as this R Notebook:

```{r}
data <- read_csv("2019_dataset_en.csv", show_col_types = FALSE)
```

## Cleaning the data
To clean the data, six columns containing unique identifiers, the year (2019),
and other variables not relevant or inhibitory to this machine learning task
must be removed as follows:

```{r}
data <- data %>%
  select(-c(C_YEAR, C_SEV, V_ID, P_ID, P_SAFE, C_CASE))
```

For the meanings of these column names, refer to the attached
"data_dictionary.pdf" for details.

The remaining columns are almost exclusively categorical variables, and contain
categories for unknown and missing data. These must be removed for the machine
learning model to process the dataset.

```{r}
data <- data %>%
  filter(C_MNTH!="UU",C_MNTH!="XX") %>%
  filter(C_WDAY!="U") %>% 
  filter(C_HOUR!="UU") %>%
  filter(C_VEHS!="UU",C_VEHS!="XX") %>%
  filter(C_CONF!="QQ",C_CONF!="UU",C_CONF!="XX") %>%
  filter(C_RCFG!="QQ",C_RCFG!="UU",C_RCFG!="XX") %>% 
  filter(C_WTHR!="U",C_WTHR!="X",C_WTHR!="Q") %>% 
  filter(C_RSUR!="U",C_RSUR!="X",C_RSUR!="Q") %>%
  filter(C_RALN!="U",C_RALN!="X",C_RALN!="Q") %>%
  filter(C_TRAF!="QQ",C_TRAF!="UU",C_TRAF!="XX") %>%
  filter(V_TYPE!="NN",V_TYPE!="QQ",V_TYPE!="UU",V_TYPE!="XX") %>%
  filter(V_YEAR!="NNNN",V_YEAR!="UUUU",V_YEAR!="XXXX") %>%
  filter(P_SEX!="N",P_SEX!="U",P_SEX!="X") %>%
  filter(P_AGE!="NN",P_AGE!="UU",P_AGE!="XX") %>%
  filter(P_PSN!="NN",P_PSN!="QQ",P_PSN!="UU",P_PSN!="XX") %>%
  filter(P_USER!="U") %>%
  filter(P_ISEV!="N",P_ISEV!="U")
```

Removing these leaves 17 of the original 23 columns and 167803 of the original
272301 observations. Finally, the most of remaining data must be converted to
factors / categorical variables, and some variables will be grouped into fewer
categories than they began with.

```{r}
# Group the C_CONF column while naming categories.
data <- data %>%
  mutate(C_CONF = (ifelse(data$C_CONF <= 10, 'Single Vehicle',
                    ifelse(data$C_CONF>20 & data$C_CONF <30, '2 Vehicles',
                    ifelse(data$C_CONF>30 & data$C_CONF<40,'2 Vehicles Head On','2 Vehicles Parked')))))

# Convert most variables to factors and some to numeric.
data$C_MNTH <- as.factor(data$C_MNTH)
data$C_WDAY <- as.factor(data$C_WDAY)
data$C_HOUR <- as.numeric(data$C_HOUR)
data$C_VEHS <- as.numeric(data$C_VEHS)
data$C_CONF <- as.factor(data$C_CONF)
data$C_RCFG <- as.factor(data$C_RCFG)
data$C_WTHR <- as.factor(data$C_WTHR)
data$C_RSUR <- as.factor(data$C_RSUR)
data$C_RALN <- as.factor(data$C_RALN)
data$C_TRAF <- as.factor(data$C_TRAF)
data$V_TYPE <- as.factor(data$V_TYPE)
data$V_YEAR <- as.numeric(data$V_YEAR)
data$P_SEX <- as.factor(data$P_SEX)
data$P_AGE <- as.numeric(data$P_AGE)
data$P_PSN <- as.factor(data$P_PSN)
data$P_ISEV <- as.factor(data$P_ISEV)
data$P_USER <- as.factor(data$P_USER)
```

To improve the readability of the data for the remainder of this script, the
values of categorical variables and the columns will be assigned English names:

```{r}
# Assign names to values of categorical variables.
data <- data %>%
  mutate(C_MNTH = recode(C_MNTH,
                        "1" = "Jan", "2" = "Feb", "3" = "Mar", "4" = "Apr",
                        "5" = "May", "6" = "Jun", "7" = "Jul", "8" = "Aug",
                        "9" = "Sep", "10" = "Oct", "11" = "Nov", "12" = "Dec")) %>%
  mutate(C_WDAY = recode(C_WDAY,
                        "1" = "Mon", "2" = "Tue", "3" = "Wed", "4" = "Thu",
                        "5" = "Fri", "6" = "Sat", "7" = "Sun")) %>%
  mutate(C_RCFG = recode(C_RCFG,
                         "1" = "Non-Intersection",
                         "2" = "Intersection - 2 public roadways",
                         "3" = "Intersection - parking, driveway etc.",
                         "4" = "Railroad level crossing",
                         "5" = "Bridge, overpass, viaduct",
                         "6" = "Tunnel or underpass",
                         "7" = "Passing or climbing lane",
                         "8" = "Ramp",
                         "9" = "Traffic circle",
                         "10" = "Express lane of a freeway",
                         "11" = "Collector lane of a freeway",
                         "12" = "Transfer lane of a freeway")) %>%
  mutate(C_WTHR = recode(C_WTHR,
                         "1" = "Clear & sunny", "2" = "Cloudy, no precipitation",
                         "3" = "Raining", "4" = "Snowing",
                         "5" = "Freezing rain, sleet, hail",
                         "6" = "Visibility limitation", "7" = "Strong wind")) %>%
  mutate(C_RSUR = recode(C_RSUR,
                         "1" = "Dry, normal", "2" = "Wet", "3" = "Snow",
                         "4" = "Slush, wet snow", "5" = "Icy", "6" = "Sand/gravel/dirt",
                         "7" = "Muddy", "8" = "Oil", "9" = "Flooded")) %>%
  mutate(C_RALN = recode(C_RALN,
                         "1" = "Straight and level", "2" = "Straight with gradient",
                         "3" = "Curved and level", "4" = "Curved with gradient",
                         "5" = "Top of hill or gradient", "6" = "bottom of hill or gradient")) %>%
  mutate(C_TRAF = recode(C_TRAF,
                         "1" = "Traffic signals fully operational",
                         "2" = "Traffic signals in flashing mode", "3" = "Stop sign",
                         "4" = "Yield sign", "5" = "Warning sign",
                         "6" = "Pedestrian crosswalk", "7" = "Police officer",
                         "8" = "School guard, flagman", "9" = "School crossing",
                         "10" = "Reduced speed zone", "11" = "No passing zone sign",
                         "12" = "Marking on the road",
                         "13" = "School bus - signal lights flashing",
                         "14" = "School bus - signal lights not flashing",
                         "15" = "Railway crossing with signals, gates",
                         "16" = "Railway crossing with signs only",
                         "17" = "Control device not specified",
                         "18" = "No control present")) %>%
  mutate(V_TYPE = recode(V_TYPE,
                         "1" = "Light duty vehicle", "5" = "Panel/cargo van",
                         "6" = "Other trucks and vans", "7" = "Unit trucks > 4536 kg",
                         "8" = "Road tractor", "9" = "School bus",
                         "10" = "Smaller school bus", "11" = "Urban and intercity bus",
                         "14" = "Motorcycle and moped", "16" = "Off road vehicles",
                         "17" = "Bicycle", "18" = "Purpose-built motorhome",
                         "19" = "Farm equipment", "20" = "Construction equipment",
                         "21" = "Fire engine", "22" = "Snowmobile", "23" = "Streetcar")) %>%
  mutate(P_PSN = recode(P_PSN,
                        "11" = "Driver", "12" = "Front row, center",
                        "13" = "Front row, right outboard", "21" = "Second row, left outboard",
                        "22" = "Second row, center", "23" = "Second Row, right outboard",
                        "31" = "Third row, left outboard", "32" = "Third row, center",
                        "33" = "Third row, right outboard", "96" = "Position unknown, but on-board",
                        "97" = "Sitting on someone's lap", "98" = "Outside passenger compartment",
                        "99" = "Pedestrian")) %>%
  mutate(P_ISEV = recode(P_ISEV,
                         "1" = "No injury", "2" = "Injury", "3" = "Fatality")) %>%
  mutate(P_USER = recode(P_USER,
                         "1" = "Motor vehicle driver", "2" = "Motor vehicle passenger",
                         "3" = "Pedestrian", "4" = "Bicyclist", "5" = "Motorcyclist"))
# Assign column names.
names(data) <- c("Month", "Weekday", "Hour", "Vehicles", "CollisionType", 
                   "RoadConfig", "Weather", "RoadSurface", 
                  "RoadAlignment", "Traffic", "VehicleType", 
                  "VehicleYear", "PersonSex", "PersonAge", 
                  "PersonPosition", "InjurySeverity", "UserType")

```

With these modifications, the data is mostly prepared for machine learning tasks,
and must now be visualized to show any more subtle problems that may appear
during training of a model.

## Exploratory Data Analysis & Visualization
First of all, getting a count of each level of injury severity reveals that the amount of collisions resulting in no injury or an injury vastly outweighs the amount of fatalities. This has implications for how we should display the data: when all 3 variables are displayed in the same graph, the fatalities are dwarfed by the no-injury and injury entries to the point where it becomes impossible to examine trends in fatalities. Therefore, we found it useful in most situations to create separate plots for fatalities in order to properly visualize them.
```{r, warning=FALSE, echo=FALSE}
summary(data$InjurySeverity)
```

### Month, weekday, and hour
```{r,out.width="150%", warning=FALSE, echo=FALSE}
# Set up histogram colors for histograms
InjurySeverity.colors <- c('Fatality' = 'brown1', 'Injury' = 'darkgoldenrod1', 'No injury' = 'darkseagreen4')

# This bit of code ensures that months display in the proper order on the bar charts
data$Month <- factor(data$Month, levels=c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec'))

#Create counts for the Month and Weekday Varaibles
Count_Month <- count(data, vars = c('Month','InjurySeverity'))
Count_Weekday <- count(data, vars = c('Weekday','InjurySeverity'))

#Month line graph
monthLine <- ggplot(Count_Month, aes(x=Month,y=freq,group=InjurySeverity,color=InjurySeverity))+
  geom_line()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_color_manual(values=InjurySeverity.colors)

#Month line graph (fatalities only)
monthLineFatal <- ggplot(Count_Month %>% filter(InjurySeverity == "Fatality"), aes(x=Month,y=freq,group=InjurySeverity,color=InjurySeverity))+
  geom_line()+
  theme(legend.position = "none")+
  ylim(0,180)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_color_manual(values=InjurySeverity.colors)

#Weekday line graph
weekdayLine <- ggplot(Count_Weekday, aes(x=Weekday,y=freq,group=InjurySeverity,color=InjurySeverity))+
  geom_line()+
  scale_color_manual(values=InjurySeverity.colors)

#Weekday line graph (fatalities only)
weekdayLineFatal <- ggplot(Count_Weekday %>% filter(InjurySeverity == "Fatality"), aes(x=Weekday,y=freq,group=InjurySeverity,color=InjurySeverity))+
  geom_line()+
  theme(legend.position = "none")+
  ylim(0,180)+
  scale_color_manual(values=InjurySeverity.colors)

#Hour density plot
hourDensity <- ggplot(data, aes(x=Hour,color=InjurySeverity))+
  geom_density()+
  scale_color_manual(values=InjurySeverity.colors)

#Display plots for day & month
monthPlots <- (monthLine|monthLineFatal)
weekdayPlots <- (weekdayLine|weekdayLineFatal)
(monthPlots/weekdayPlots) + plot_layout(guides = "collect")

#Display hour density plot
hourDensity
```
With these graphs, we can see that collisions of all types follow the same general trends over the course of the year & course of each week. The shape of the lines for No Injury and Injury are nearly the same in both graphs. Here are some trends of note:

 - More collisions in summer months, peaking in June
 - Less collisions in winter months, bottoming out in April
 - Collisions dip in September, then rise again in October before resuming the downwards trend
 
Fatalities follow the same general pattern, but note that there seems to be a consistent low level of fatalities from Jan-Apr, whereas for other types of collisions the number starts higher and then decreases.

Over the course of a week, overall collisions: 

 - Increase over the week, peaking on Friday. 
 - Decrease sharply over the weekend
 
The decrease over the weekend is likely due to reduced commuting levels. Fatalities follow the same general pattern, but are slightly higher on Mondays & Tuesdays, with the lowest amount of fatalities on Wednesday.

The density plot of collisions over the course of the day shows that collisions are likely correlated by overall commuting patterns. Here are some observations:

 - Overall collisions have a small spike at about 7am (hour 7)
 - Larger spike after about 3pm (hour 15)
 - Collisions are lowest between midnight and 5am (hours 0-5) 
 - Comparatively, fatalities go down less between hours 0 and 5. This could be due to decreased visibility at night.

### Demographics
```{r,out.width="150%", warning=FALSE, echo=FALSE}
#PersonSex bar plots
sexBar <- ggplot(data, aes(x=PersonSex,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  scale_fill_manual(values=InjurySeverity.colors)

sexBarFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=PersonSex,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  theme(legend.position = "none")+
  scale_fill_manual(values=InjurySeverity.colors)

#PersonAge density plot
ageDensity <- ggplot(data, aes(x=PersonAge,color=InjurySeverity))+
  geom_density()+
  scale_color_manual(values=InjurySeverity.colors)

#Plots for demographics
sexPlots <- sexBar + sexBarFatal 
(sexPlots) + plot_layout(guides = "collect")

ageDensity
```
Plotting the data for sex reveals some insights:
 - There are more than 2 times as many male deaths compared to female deaths
 - Collisions involving women have a higher proportion of injuries to no injuries
 - Collisions involving men have a slightly higher rate of no injuries to injuries

In the density plot for injury vs age: 
  - No Injury and Injury rates are nearly identical, except in children under 12, who have a lower rate of injury
  - Children under the age of 12 also have a lower share of deaths.
  - Over the age of about 50, there are proportionately more deaths

### Road & Weather Conditions
```{r, warning=FALSE, echo=FALSE}
#Weather
weatherBar <- ggplot(data, aes(x=Weather,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#RoadSurface
roadSurfaceBar <- ggplot(data, aes(x=RoadSurface,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

weatherRoadCondBars <- weatherBar + roadSurfaceBar & theme(legend.position = "right")
(weatherRoadCondBars) + plot_layout(guides = "collect")
```
Making an initial set of bar charts for weather & road surface conditions, we can see a problem right away - an overwhelming majority of these entries were recorded with clear/sunny weather and dry/normal road conditions. To some this might imply that vehicle collisions are much more likely to happen under clear & dry conditions, but intuitively we know this isn't true. Instead it likely means that there was mostly sunny & dry conditions at most times throughout the country. Also, we are not trying to find out under which conditions someone is likely to have a vehicle collision - we are trying to find out which conditions result in a higher chance of injury or fatality if one gets in a vehicle collision.

The charts can be redone to show proportions: 

```{r, warning=FALSE, echo=FALSE}
#Count_Weather <- count(data, vars = c('Weather','InjurySeverity'))
#Count_RoadSurface <- count(data, vars = c('RoadSurface','InjurySeverity'))

#Weather
weatherBar_proportion <- ggplot(data, aes(x=Weather, fill=InjurySeverity, by=Weather))+
  geom_bar(position="fill")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#RoadSurface
roadSurfaceBar_proportion <- ggplot(data, aes(x=RoadSurface, fill=InjurySeverity, by=RoadSurface))+
  geom_bar(position="fill")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

weatherRoadCondBars_proportion <- weatherBar_proportion + roadSurfaceBar_proportion & theme(legend.position = "right")
(weatherRoadCondBars_proportion) + plot_layout(guides = "collect")
```


```{r, warning=FALSE, echo=FALSE}
#Number of vehicles involved
numVehiclesHist <- ggplot(data, aes(x=Vehicles,fill=InjurySeverity))+
  geom_bar(bins=24, position="dodge")+
  scale_fill_manual(values=InjurySeverity.colors)

numVehiclesHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Vehicles,fill=InjurySeverity))+
  geom_bar(bins=24)+
  scale_fill_manual(values=InjurySeverity.colors)

#RoadConfig
roadConfigHist <- ggplot(data, aes(x=RoadConfig,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

roadConfigHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=RoadConfig,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)



weatherHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Weather,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)



roadSurfaceHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=RoadSurface,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#RoadAlignment
roadAlignmentHist <- ggplot(data, aes(x=RoadAlignment,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

roadAlignmentHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=RoadAlignment,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#Traffic
trafficHist <- ggplot(data, aes(x=Traffic,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

trafficHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=Traffic,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#VehicleType
vehTypeDensity <- ggplot(data, aes(x=VehicleType,color=InjurySeverity))+
  geom_density(stat="count", position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_color_manual(values=InjurySeverity.colors)

vehTypeHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=VehicleType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#VehicleYear: density plot
vehYearDensity <- ggplot(data %>% filter(VehicleYear > 1975), aes(x=VehicleYear,color=InjurySeverity))+
  geom_density(position="dodge")+
  scale_color_manual(values=InjurySeverity.colors)

vehYearHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=VehicleYear,fill=InjurySeverity))+
  geom_bar(bins=100)+
  scale_fill_manual(values=InjurySeverity.colors)





#PersonPosition
pPositionHist <- ggplot(data, aes(x=PersonPosition,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

pPositionHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=PersonPosition,fill=InjurySeverity))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=InjurySeverity.colors)

#UserType
pUserTypeHist <- ggplot(data, aes(x=UserType,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  scale_fill_manual(values=InjurySeverity.colors)

pUserTypeHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=UserType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)

#collisionType
collisionTypeHist <- ggplot(data, aes(x=CollisionType,fill=InjurySeverity))+
  geom_bar(stat="count", position="dodge")+
  scale_fill_manual(values=InjurySeverity.colors)

collisionTypeHistFatal <- ggplot(data %>% filter(InjurySeverity == "Fatality"), aes(x=CollisionType,fill=InjurySeverity))+
  geom_bar(stat="count")+
  scale_fill_manual(values=InjurySeverity.colors)
```

First, we create bar charts to see how many collisions of each severity there are according to each variable. Most of the variables are either categorical, or integer numbers so bar charts are the best option.

Bar charts are displayed grouped by theme:

 - Time
 - Demographics
 - Weather & road conditions
 - Vehicle info
 - Road info
 - Driver/passenger info
 - Collision info
 
 Note that for all variables, the amount of fatalities is drastically lower than the amount of injuries and non-injuries, to the point where it's impossible to view patterns in the data. Therefore, for each varaible there is a separate chart to show how fatalities correlate with each variable.
 
 
 A few observations can be made with these histograms. To start, the histogram displaying the number of vehicles is drastically right skewed. Most collisions that occur involve at most 5 vehicles with the majority only involving 2. 
 
The road configuration histogram shows us that there are primarily two cases of configuration consisting of either non-intersection or intersections at 2 roadways. The other cases are very uniquely rare conditions, which we see a few times in the data. The interesting observation here is that the proportion of fatalities at non-intersections are much higher than intersections. 

Road alignment suffers from primarily a single outcome of straight and level roads. There is a noticeable impact however of curved and level surfaces involving fatalities. 

There are 3 impactful factors for traffic control, namely traffic signals not fully operational, stop sign, or primarily no control present. This is consistent with fatality cases. 

Vehicle type accidents are primarily classed as light duty vehicles as these are used most often for daily commutes. It should be noted that motorcycles are impactful when observing fatalities, 

The histogram for vehicle year is left skewed, and most years rest within 2012 to 2020. When switching to fatalities, the dates seem to shift towards older vehicles, indicating what could be a lack in safety features and overall condition of the infrastructure of older vehicles. 

This spike in collision cases across all severities after the age of 16 is likely a result of inexperienced drivers and as age increases the driving experience would as well. 
 
 

```{r, out.width="150%", warnings=FALSE, echo=FALSE}
#Display similar grouped histograms:





#Histograms for road & weather conditions


#Histograms for vehicle info
vehicleInfoHists <- vehYearDensity & theme(legend.position = "right")
#vehicleInfoHistsFatal <- vehTypeHistFatal + vehYearHistFatal & theme(legend.position = "none")
vehicleInfoHists + plot_layout(guides = "collect")

#Histograms for road info
roadInfoHists <- roadAlignmentHist + trafficHist + roadConfigHist & theme(legend.position = "right")
roadInfoHistsFatal <- roadAlignmentHistFatal + trafficHistFatal + roadConfigHistFatal & theme(legend.position = "none")
(roadInfoHists/roadInfoHistsFatal) + plot_layout(guides = "collect")

#Histograms for driver/passenger info
driverPassengerHists <- pPositionHist / pUserTypeHist & theme(legend.position = "right")
driverPassengerHistsFatal <- pPositionHistFatal + pUserTypeHistFatal & theme(legend.position = "none")
(driverPassengerHists/driverPassengerHistsFatal) + plot_layout(guides = "collect")

#Histograms for collision info
collisionHists <- collisionTypeHist + numVehiclesHist & theme(legend.position = "right")
collisionHistsFatal <- collisionTypeHistFatal + numVehiclesHistFatal & theme(legend.position = "none")
(collisionHists/collisionHistsFatal) + plot_layout(guides = "collect")
```


The data is distributed in such a way that the majority of the results of the target variable 
InjurySeverity are either No injury or Injury. No injury makes up (76530)/167803 or 45.6% of cases, 
Injury makes up (90299)/167807 or 53.8% leaving only 974/167807 or 0.5% resulting in fatalities.  
Fortunately, given the size of the data, we can acquire 3 suitably equal sized samples of the data 
for training and testing purposes.


```{r}
fatalSet <- data %>%
  filter(InjurySeverity == "Fatality") %>%
  sample_n(900)

injurySet <- data %>%
  filter(InjurySeverity == "Injury") %>%
  sample_n(900)

noInjurySet <- data %>%
  filter(InjurySeverity == "No injury") %>%
  sample_n((900))

data <- bind_rows(fatalSet,injurySet,noInjurySet, .id = NULL)

```



A partition is created to create separate subsets of our data for the purpose of training and validating the models.
75% of our set will be used to train the data, and the remaining 25% will be used to validate the models we choose.

```{r}
inTraining <- createDataPartition(data$InjurySeverity, p=0.75, list=FALSE)
training <- data[inTraining,]
validation <- data[-inTraining,]
```


Two parameters will be needed for the models. One is the method in which the control data will be training the model. The method used is cross-validation. This method splits the dataset randomly into groups and runs a series of tests that store results used to measure the metric of interest. We will repeat this for 20 unique tests sets. The metric of interest is the accuracy of the model.

```{r}
control <- trainControl(method="cv", number=20)
metric <- "Accuracy"
```


Running the Models:

LDA (Linear Discriminant Analysis):

This model works well with categorical predictor variables, however requires some assumptions be met. If the predictors are not normally distributed, could cause problems.






C&RT (Classification and Regression Trees):

This model excels when the data consists of observations broken into groups. There are a lot of categorical variables with multiple levels of factors making this model an ideal choice at first glance.


```{r}
cartModel <- train(InjurySeverity~., data=training, method="rpart", metric=metric, trControl=control)
predictions <- predict(cartModel, validation)

cmCart <- confusionMatrix(predictions, as.factor(validation$InjurySeverity))
cmCart

summary(cmCart)
```



KNN (K-Nearest Neighbors):

```{r}
knnModel <- train(InjurySeverity~., data=training, method="knn", metric=metric, trControl=control)
predictions <- predict(knnModel, validation)

cmKnn <- confusionMatrix(predictions, as.factor(validation$InjurySeverity))
cmKnn
```



SVM (Support Vector Machine):

```{r}
svmModel <- train(InjurySeverity~., data=training, method="svmRadial", metric=metric, trControl=control)
predictions <- predict(svmModel, validation)

cmSvm <- confusionMatrix(predictions, as.factor(validation$InjurySeverity))
cmSvm
```


